Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-02-14 infercnv/reports/gains_and_losses.Rmd
looking for sub-chromosome-level copy number events, add nebulosa density plotting for umap, change gains_and_losses pseudobulk to all cells

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  infercnv_dir: NULL
  cache_dir: NULL
  rerun: true
---

@@

# Pre-processing

@@

First, we load the CITE-seq object. We convert it to a SingleCellExperiment object. Next, we add the CNV metadata from the inferCNV run to the object.

```{r load_obj, results = 'asis'}
# dirs
data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
infercnv_dir <- params$infercnv_dir

@@

@@

  # get cnv metadata
  cnvs <- 
    list.files(
      infercnv_dir,
      pattern = 'map_metadata_from_infercnv.txt',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(infercnv_dir, file) %>% 
        read.table(sep = '\t')
      }) %>%
    dplyr::bind_rows()

  # left join cnvs to coldata, reorder, restore rownames
  col_data <- merge(col_data, cnvs, by = 'row.names', all.x = T)
  rownames(col_data) <- col_data[, 1]
  col_data <- 
    col_data[colnames(sce), -1] %>%
    DataFrame()

@@

}, file = 'sce.rds', rerun = params$rerun)
```

@@

First, we plot UMAPs by ID and celltype.

@@

Then, we visualise chr7 gains on the UMAP.

```{r plot_chr_gains}
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr7', size = 0.1) +
  coord_fixed()
```

Now, we will plot gain and loss events by chromosome.

@@

```{r plot_gains_and_losses_by_ct}
# at single-cell level
unique(sce$cell_type) %>%
  purrr::map(function(ct) {
    c('dupli', 'loss') %>%
      purrr::map(function(direction) {
        p_dat <-
          sce@colData[!is.na(sce$proportion_cnv_chr1) & sce$cell_type == ct, ]
        p_mat <-
          p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
        p_annot <- 
          p_dat[, c('sample'), drop = F] %>%
          as.data.frame()
        pheatmap::pheatmap(
          p_mat, 
          show_rownames = F, 
          annotation_row = p_annot,
          # hclust error: size cannot be NA nor exceed 65536
          cluster_rows = ifelse(nrow(p_mat) > 65536, F, T),
          legend = F,
          main = paste0('celltype: ', ct, '\ndirection: ', direction)) %>%
          print()
      })
  })
```

@@

# chr7 gain

@@

  # pseudobulk the sce object by patient and chr7 ploidy
  # subset to stromal cells
  reduced_sce <-
    glmGamPoi::pseudobulk(
      sce[, sce$cell_type == 'Stromal cell' & !is.na(sce$chr7_gain)], 
      group_by = vars(sample, chr7_gain)
    )

  # remove NA's
  reduced_sce <-
    reduced_sce[, !is.na(reduced_sce$sample) & !is.na(reduced_sce$chr7_gain)]

@@

```{r deseq, message = T}
# fit one GLM model for each gene
fit <- xfun::cache_rds({
  fit <-
  glmGamPoi::glm_gp(
    reduced_sce,
    design = ~ chr7_gain + sample,
    verbose = T
  )
  fit
}, file = 'fit.rds', rerun = params$rerun)

@@

  # test differences
  de_res <-
    glmGamPoi::test_de(
      fit, 
      contrast = 
        cond(chr7_gain = 1) -
        cond(chr7_gain = 0)
    )

>>>>

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  infercnv_dir: NULL
  cache_dir: NULL
  rerun: true
---

@@

# Loading the inferCNV outputs

@@

First, we load the CITE-seq object. We convert it to a SingleCellExperiment object. Then we add the CNV metadata from the inferCNV run to the object.

```{r load_obj, results = 'asis'}
# dirs
data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'

@@

  # load CTAP and metadata
  ctap <-
    readr::read_tsv('data/Zhang2023/CTAP_donor_mapping.tsv') %>%
    dplyr::select(sample = donor, CTAP)
  md <- 
    readr::read_tsv('data/Zhang2023/SYNAPSE_METADATA_MANIFEST.tsv') %>%
    dplyr::distinct()

@@

  # get cnv metadata
  cnvs <- 
    list.files(
      params$infercnv_dir,
      pattern = 'map_metadata_from_infercnv.txt',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(params$infercnv_dir, file) %>% 
        read.table(sep = '\t')
      }) %>%
    dplyr::bind_rows() %>%
    dplyr::mutate(infercnv = T)

  # left join cnvs to coldata, reorder, restore rownames
  col_data <- merge(col_data, cnvs, by = 'row.names', all.x = T, sort = F)
  rownames(col_data) <- col_data[, 1]
  col_data <- 
    col_data[colnames(sce), -1] %>%
    DataFrame()

@@

}, file = 'sce.rds', rerun = params$rerun)

 # remove NA's for inferCNV analysis
 sce <- sce[, !is.na(sce$proportion_cnv_chr1)]
```

## `map_metadata_from_infercnv.txt`

@@

## HMM regional predictions

We also load the CNV regions prediced by HMM. According to [their wiki](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type#infercnv-i6-hmm)...

As well as aneuploidy, we want to look for smaller recurrent structural changes, including focal-/arm-level events.

> The inferCNV i6 HMM is a six-state CNV model that predicts the following CNV levels:
>
> * State 1: 0x: complete loss
> * State 2: 0.5x: loss of one copy
> * State 3: 1x: neutral
> * State 4: 1.5x: addition of one copy
> * State 5: 2x: addition of two copies
> * State 6: 3x: essentially a placeholder for >2x copies but modeled as 3x.

```{r get_hmm}
cnv_regions <- xfun::cache_rds({
  cnv_regions <- 
    list.files(
      params$infercnv_dir,
      pattern = 'HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(params$infercnv_dir, file) %>% 
        readr::read_tsv()
      }) %>%
    dplyr::bind_rows(.id = 'sample') %>%
    # refactor chromosomes
    dplyr::mutate(
      chr = chr %>% {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
      direction = dplyr::case_when(
        state < 3 ~ 'loss',
        state > 3 ~ 'dupli',
        TRUE ~ 'neutral'),
      region = gsub('.*region_', '', cnv_name) %>% as.numeric())
}, file = 'cnv_regions', rerun = params$rerun)
```

```{r plot_cnv_regions_pred}
pdf('test.pdf', height = 25)
cnv_regions %>%
  dplyr::filter(direction != 'neutral') %>% 
  ggplot(aes(x = region)) +
  geom_bar(position = 'stack') +
  facet_wrap(~ chr, ncol = 1, scales = 'free') +
  theme_classic() +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme(legend.position = 'none')
dev.off()
```

## UMAPs

First, we plot UMAPs by celltype.

@@

Then, we visualise gains and loss proportions for each chromosome on the UMAP.

```{r plot_chr_gains, fig.width = 10, fig.height = 5}
pdf('test.pdf')
paste0('chr', 1:22) %>% head(2) %>%
  purrr::map(function(chr) {
    sce %>%
      Nebulosa::plot_density(
        paste('proportion', c('dupli', 'loss'), chr, sep = '_'),
        size = 0.5) 
  })
```

## Heatmaps

Now, we will plot gain and loss proportions by chromosome.

@@

```{r plot_gains_and_losses_by_ct}
# at single-cell level
sce$cell_type[!is.na(sce$infercnv)] %>%
  unique() %>%
  purrr::map(function(ct) {
    c('dupli', 'loss') %>%
      purrr::map(function(direction) {
        p_dat <-
          sce@colData[!is.na(sce$proportion_cnv_chr1) & sce$cell_type == ct, ]
        if (nrow(p_dat) > 0) {
          p_mat <-
            p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
          p_annot <- 
            p_dat[, c('sample'), drop = F] %>%
            as.data.frame()
          pheatmap::pheatmap(
            p_mat, 
            show_rownames = F, 
            annotation_row = p_annot,
            # hclust error: size cannot be NA nor exceed 65536
            cluster_rows = ifelse(nrow(p_mat) > 65536, F, T),
            legend = F,
            main = paste0('celltype: ', ct, '\ndirection: ', direction)) %>%
            print()
        }
      })
  })
```

@@

## chr7 gain

@@

  # pseudobulk the sce object by patient and chr7 ploidy
  # subset to stromal cells
  reduced_sce <-
    glmGamPoi::pseudobulk(
      sce[, !is.na(sce$sample) & 
            !is.na(sce$chr7_gain) &
            !is.na(sce$cell_type)], 
      group_by = vars(sample, cell_type, chr7_gain)
    )

@@

```{r deseq, message = T}
# fit one GLM model for each gene
fit <- xfun::cache_rds({
  fit <-
  glmGamPoi::glm_gp(
    reduced_sce,
    design = ~ chr7_gain + sample + cell_type,
    verbose = T
  )
  fit
}, file = 'fit.rds', rerun = params$rerun)

@@

  # test differences
  de_res <-
    glmGamPoi::test_de(
      fit, 
      contrast = 
        cond(chr7_gain = 1, cell_type = 'Stromal cells') -
        cond(chr7_gain = 0, cell_type = 'Stromal cells')
    )
